
## Solution to exercise 1

3.  Print your working directory

First, you had to print your working directory. To do this in quarto, you have to add a new cunk code, which you can do by option + command + i in MAC or ... in Windocs. You can also click on code and 'Insert Chunk' in the functions bar.

```{r}
getwd() 
here::here() 
```

The `haven` package allows you to read both files from Stata or from SPSS with either `read_dta()` or `read_sav()`. Here I've copied the address of the pdf link on the site and read the file directly from the Internet. I could also have downloaded the file and saved it in the same folder as my project to import it, specifying the file name as the first argument to the `read_dta` function.

```{r}
qog <- haven::read_dta("https://www.qogdata.pol.gu.se/data/qog_ei_ts_sept21.dta")
```

```{r}
qog |> head(10) # Look at the first 10 rows
qog |> tail(5) # Look at the last 5 rows
dplyr::glimpse(qog)
```

Or we can have a look at the codebook which we can download with the data that gives us a description of every variable in the dataset.

```{r}
summary(qog$wdi_fossil)
mean(qog$wdi_fossil)
mean(qog$wdi_fossil, na.rm = TRUE)
```
