# Exercise 1 {.unnumbered}

1.  Create an R project on your computer in the location and with the name you want (avoid white spaces). This should create a new folder containing an Rproj file.

2.  Create a quarto document with the title "Exercise 1" and your name as author. Save the file as ("lastname_ex1.qmd") and save it in the same folder : https://r4ds.hadley.nz/quarto.html

3.  Add a chunk code and print out your working directory to make sure your R session is launched in the project you've created. The last part of the path displayed should correspond to the project name. ATTENTION : POUR L'INSTANT : add working directory explanations.

4.  Add a new chunk code and find a way to import the [Quality of Government environmental indicators](https://www.gu.se/en/quality-government/qog-data/data-downloads/environmental-indicators-dataset) dataset into Rstudio, either in `.dta` format (stata) or `.sav` format (spss).

5.  Once the data has been imported, answer the following questions with both code and writing on the quarto document:

-   What is this dataset about ?
-   What is the unit of analysis for this dataset ?
-   Find a way to calculate the mean of the variable wdi_fossil and write a sentence to interpret the result.

6.  Render the quarto document in a pdf format and upload your file on moodle. To compile in pdf, you might need to install the `tinytext` package and it might actually do not work. If it does not just compile into html and send it like this. If you do not succeed some of the exercice and render fail because of errors, add `#| error:true` in the chunk code that produces errors. It will allow me to see your errors.

## Solution to exercise 1

3.  Print your working directory

First, you had to print your working directory. To do this in quarto, you have to add a new cunk code, which you can do by option + command + i in MAC or ... in Windocs. You can also click on code and 'Insert Chunk' in the functions bar.

```{r}
getwd() 
here::here() 
```

The `haven` package allows you to read both files from Stata or from SPSS with either `read_dta()` or `read_sav()`. Here I've copied the address of the pdf link on the site and read the file directly from the Internet. I could also have downloaded the file and saved it in the same folder as my project to import it, specifying the file name as the first argument to the `read_dta` function.

```{r}
qog <- haven::read_dta("https://www.qogdata.pol.gu.se/data/qog_ei_ts_sept21.dta")
```

```{r}
qog |> head(10) # Look at the first 10 rows
qog |> tail(5) # Look at the last 5 rows
dplyr::glimpse(qog)
```

Or we can have a look at the codebook which we can download with the data that gives us a description of every variable in the dataset.

```{r}
summary(qog$wdi_fossil)
mean(qog$wdi_fossil)
mean(qog$wdi_fossil, na.rm = TRUE)
```
